Tool-driven Revolutions in Archaeological Science

### 2019-08-08 analysis/paper/jcaa-paper.Rmd
up to L470 of reviewer B's edits

As @Galison_1997 describes a shift in the practices of physicists, we believe so is archaeological practice changing in the digital age. @Huvila_2018 have explored the use of digital tools on archaeological practices to understand the boundaries of what counts as archaeological, what is merely related to archaeology and what should be excluded from archaeology. Reflecting on @Huvila_2018, our view is the use use of code is unquestionably archaeology, since it is just one of many tools that archaeologists are using for converting raw data into summaries, visualisations, and insights. These concerns about how digital tools affect what we consider is and is not archaeology intersect with the work of @gieryn1999cultural on boundary construction. In his cultural cartography of science, Gieryn examines boundaries between science generally and varieties of less authoritative non-science. He argues that the boundaries are drawn and redrawn contiuously in flexible, historically changing and sometimes ambiguous ways. Gieryn claims that the cultural authority of science flows from boundary work in professional and political settings where scientists construct a public image for science as a source of credible knowledge, validate their work, and marginalize claims by competitors. Boundary work involves attributions of select characteristics to science in order to distinguish it from non-scientific competitors. The increasing voluntary use and sharing of code by researchers, and demands for it by readers and peer reviewers is a form of active boundary work that is shifting the boundary between scientific and non- or less-scientific archaeology. We, as members of this group, argue that the use of code as a research tool, and the sharing of code as a public research product, are increasingly attributes that are being used to distinguish science from less scientific work. The reason for this is that using code, and making it available for others to inspect, enhances the credibility of knowledge claims and the visibility of those claims.

A related issue here is the _technicity_ of tools, in our case coding. We draw on @cassirer_form_1930 and @simondon_mode_1958 to define technicity as a term to describe the technology, i.e. tools and machines respectively, that interact with humans and with nature [@hoel_ontological_2013, 188]. Technology is said to own an instrumental kind of logos (Cassierer, see @hoel_ontological_2013, 194) and in analogy to language it exercises a certain measure of agency in mediating between humans, nature and technology. The emphasis here is on the function of technology, in the forces that it exercises, with a certain measure of autonomy, on other beings as well as in and through the new realities, it brings into being. The use of a scripting language might exemplify this, as it is not just a technological tool, but a proper "language" as well. Scripts of code are texts that translate data into visualisations and statements about the world, transversing and revitalizing conceptions of the relationship between people and nature, and between people as members of a research community.

The concept of technicity is relevant here because in contrast to the concept of technological determinism, which puts technology as a force against the human, van der Tuin and Hoel (as well as Cassirer and Simondon) enfold technology into 'the human' without letting it lose its 'foreignness' and further develop the notions of 'human' and 'nature' [@hoel_ontological_2013, 190]. Any change in tools therefore will influence how human researchers interact with their objects of interest and other people in their research community. For example, tools that make the research process more transparent and open should make the research community more open and inclusive to diverse participants and ideas. Using code to 'stand for' or represent the research process will increase our awareness of how archaeology itself 'stands for' the human experience in the past, and how it is a representation of the past [cf. @pearson1998beginning]. Similarly, the code has to be successfully executed to generate a useful result, just as claims about the past have to be successfully engaged with archaeological evidence to count as plausible.  Interestingly it is exactly at the point between continously evolving subjects and objects, where van der Tuin and Hoel place technology and the creation of knowledge [@hoel_ontological_2013, 199]. The "productive back-and-forth" between all parties may challenge established notions of culture, but also offers a concept of formation through differing, called "performative correspondence" [@hoel_ontological_2013, 201]. The potential of a certain technology cannot be known before its coming-into-being as it will reflect back on its creation ("concretization" in Simondon's terms) as well as interact with other technologies (developed or yet-to-be developed) [@hoel_ontological_2013, 194]. Observing how scientific scripting languages have become a concrete tool in ecology leads us to expect predict similar trends in archaeology.

>>>>

As @Galison_1997 describes a shift in the practices of physicists, we believe archaeological practice is similarly changing. @Huvila_2018 have explored the use of digital tools on archaeological practices to understand the boundaries of what counts as archaeological, what is merely related to archaeology and what should be excluded from archaeology. Reflecting on @Huvila_2018, the use of code is unquestionably archaeological, since it is one of many tools that archaeologists are using for converting raw data into summaries, visualisations, and insights. These concerns about how digital tools affect what we consider is and is not archaeology intersect with the work of @gieryn1999cultural on boundary construction. In his cultural cartography of science, Gieryn examines boundaries between science generally and varieties of less authoritative non-science. He argues that the boundaries are drawn and redrawn contiuously in flexible and sometimes ambiguous ways. Gieryn claims that the cultural authority of science flows from boundary work in professional and political settings where scientists construct a public image for science as a source of credible knowledge, validate their work, and marginalize claims made by their competitors in the scientific community. Boundary work involves attributions of select characteristics to science in order to distinguish it from non-scientific competitors. The increasing voluntary use and sharing of code by researchers, and demands for it by readers and peer reviewers is a form of active boundary work that is shifting the boundary between scientific and non- or less-scientific archaeology. We, as members of this group, argue that the use of code as a research tool, and the sharing of code as a public research product, are increasingly attributes that are being used to distinguish science from less- and un-scientific work. The reason for this is that using code, and making it available for others to inspect, enhances the credibility of knowledge claims and the visibility of those claims.

A related issue here is the _technicity_ of tools, in our case programming languages @cassirer_form_1930 and @simondon_mode_1958 to define technicity as a term to describe the technology, i.e. tools and machines respectively, that interact with humans and with nature [@hoel_ontological_2013, 188]. Technology is said to own an instrumental kind of logos (Cassierer, see @hoel_ontological_2013, 194) and in analogy to language it exercises a certain measure of agency in mediating between humans, nature and technology. The use of a scripting language might exemplify this, as it is not just a technological tool, but a proper "language" as well. Scripts of code are texts that translate data into visualisations and statements about the world, transversing and revitalizing conceptions of the relationship between people and nature, and between people as members of a research community.

In contrast to the concept of technological determinism, which puts technology as a force against the human experience, technicity enfolds technology into 'the human' without letting it lose its 'foreignness' [@hoel_ontological_2013, 190]. According to this perspective, changes in tools will influence how human researchers interact with their objects of interest and other people in their research community. There are three ways is which this can be anticipated for archaeology. Firstly, tools that make the research process more transparent and open might be expected to make the research community more open and inclusive to diverse participants and ideas. Secondly, using code to 'stand for' or represent the research process will likely increase our awareness of how archaeology itself 'stands for' the human experience in the past, and how it is a representation of the past [cf. @pearson1998beginning]. Thirdly, the code has to be successfully executed to generate a useful result, just as claims about the past have to be successfully engaged with archaeological evidence to count as plausible.
